Conversational expectations account for apparent limits on theory of mind use

### 2017-09-19 writing/journal_manuscript/pnas-submission.tex
Minor edits to intro

%While these failures are compelling evidence against a simple model where theory of mind reasoning is limited to literal perspective-taking, 
We argue in this paper that the mental models supporting theory of mind use in communication are much richer than past efforts have explored. 
We consider a class of mental models representing not only a partner's visual access but also Gricean expectations of rationality and cooperativity \cite{Grice75_LogicConversation}: that speakers will avoid saying things that are confusing or unnecessarily complicated given the current context. 
For example, suppose a friend wants to refer to a nearby Dalmatian. 
%There are numerous descriptions that would all be literally true, but different expressions are preferred in different contexts.
If there are no other dogs nearby, we may expect them to use a parsimonious basic-level term like ``dog.''
When there is potential for confusion, however, we expect them to introduce additional modifiers (``spotted dog'') or shift to more specific terms (``Dalmatian'') to disambiguate the one they are trying to refer to \cite{BrennanClark96_ConceptualPactsConversation, GrafEtAl16_BasicLevel}. When speakers violate these expectations, listeners draw inferences about the intended meaning in order to account for the violation: for example, XXX \cite{XXX, BrehenyKatsosWilliams06_ImplicaturesOnline, StillerGoodmanFrank15_AdHocImplicature}.

@@

Theory of mind mechanisms are therefore at the very heart of the Rational Speech Act framework, yet they have never been directly compared against egocentric or ``mindblind'' approaches under the same behavioral paradigm. By formalizing and rigorously comparing the predicted behavior of speakers and listeners using mental models of one another, we show that theory of mind is required to account for context-sensitive behavior. In addition to these novel predictions, this approach also allows us to understand the sophisticated pragmatics underlying behavior that, under more naive behavioral measures, apparently reflect failures of theory of mind. %Communication abounds with potential social expectations, and evaluating the extent to which theory of mind is deployed in communication requires considering these expectations \emph{en masse}. 
While egocentric views of perspective-taking make more clear-cut behavioral predictions, the predictions derived from a computational theory of mind are more varied and complicated but ultimately more accurate.

>>>>

%While these failures are compelling evidence against a simple model where theory of mind reasoning is limited to literal perspective-taking, 
We argue in this paper that the mental models supporting theory of mind use in communication are much richer than past efforts have explored. 
We consider a class of mental models representing not only a partner's visual access but also Gricean expectations of parsimony and cooperativity \cite{Grice75_LogicConversation}: that speakers will avoid saying things that are confusing or unnecessarily complicated given the current context. 
For example, suppose a friend wants to refer to a nearby Dalmatian. 
%There are numerous descriptions that would all be literally true, but different expressions are preferred in different contexts.
If there are no other dogs nearby, we may expect them to use a parsimonious basic-level term like ``dog.''
When there is potential for confusion, however, we expect them to introduce additional modifiers (``spotted dog'') or shift to more specific terms (``Dalmatian'') to disambiguate the one they are trying to refer to \cite{BrennanClark96_ConceptualPactsConversation, GrafEtAl16_BasicLevel}. When speakers violate these expectations, listeners draw inferences about the intended meaning in order to account for the violation: for example, XXX \cite{XXX, BrehenyKatsosWilliams06_ImplicaturesOnline, StillerGoodmanFrank15_AdHocImplicature}.

@@

%Since positive claims about theory of mind in communication concern the mental model we hold of our partner, these claims are best evaluated in a computational framework that formalizes the content of our representations and how we use them. We test a richer, more realistic theory of mind accounting not just for visual access but also for Gricean expectations of parsimony and context-appropriate informativity under uncertainty. Under this richer model, we find that behavior taken as evidence for limits on theory of mind reasoning is in fact precisely what is required by the pragmatics of the situation.

Theory of mind mechanisms are therefore at the heart of the Rational Speech Act framework, yet they have never been directly compared against the predictions of egocentric or ``mindblind'' theories under the same behavioral paradigm. Since positive claims about theory of mind in communication concern the mental model we hold of our partner, these claims are best evaluated in a computational framework that formalizes the content of our representations and how we use them. By rigorously comparing the predicted behavior of speakers and listeners using different mental models, we show that theory of mind is required to account for context-sensitive behavior. In addition, we replicate effects that, under more naive behavioral measures, were interpreted as evidence of limits on theory of mind use, but under a richer computational model are shown to be rational consequences of an uncooperative confederate. %Communication abounds with potential social expectations, and evaluating the extent to which theory of mind is deployed in communication requires considering these expectations \emph{en masse}. 
While egocentric views of perspective-taking make more clear-cut behavioral predictions, the predictions derived from a computational theory of mind are more varied and complicated but ultimately more accurate.
